The LSST-DESC 3x2pt Tomography Optimization Challenge

### 2021-03-25 paper/paper.tex
more malzing

% AIM: messaging -- establish scope with the basic info now in the abstract
This paper is part of preparations for the analysis of Legacy Survey of Space and Time (LSST) data at the Rubin Observatory, being run by the Dark Energy Science Collaboration (DESC).
In it, we discuss and evaluate the related challenge
of using a limited set of colour bands to make those assignments, motivated by requirements of using the
metacalibration method for galaxy shape measurement to limit biases.  
A similar question, with different motivations, was explored in \citet{jain}, who find that 3-band 
gri tomography could be effective for tomographic selection, a result we will echo here in a new context: in this paper we describe the results of a 
challenge using simulated Rubin-like data with a range of methods submitted by different teams.  We explore how many tomographic bins we can effectively generate using only a subset of bins, and compare methods submitted to the challenge.

@@

In the challenge, participants assigned simulated galaxies to tomographic bins,
using only their (g)riz bands and associated errors, and the galay size.  Their goal was to maximize the cosmological
information in the final split data set; this is generally acheived by separating different bins by 
redshift as much as possible.  Particpants were free to optimize or simply select nominal (target) 
redshift bin edges from their training sample, but in either case had to classify galaxies into the 
different bins.

@@

The challenge was structured as a python library, in which each classification
method was a subclass of an abstract base parent superclass, {\sc Tomographer}.
The superclass, and other challenge machinery, performed initialization,
configuration option handling, data loading, and computing derived quantities
such as colours from magnitudes.

@@

\subsection{Results Overview}
% AIM: Suggestion: Convert this subsection to a bulleted list to give it some structure harking to the detailed subsections to come. The short versions are long enough to come off as unsupported statements rather than previews of what will be demonstrated in subsequent subsections.

The DETF results for each method are shown in Table \ref{tab:cosmodc2}.  Complete results for all metrics
for the nine-bin scenario only are shown in appendix tables \ref{tab:full_dc2} and \ref{tab:full_buzz}.  
% AIM: Use autoref to appendix.
In all these tables, missing entries are shown for two reasons.
The first, marked with an asterisk, is when a method did not run for the given scenario,
either due to memory overflow or taking an extremely long time.
The second, marked with a dash, is when a method ran, but generated at least one bin with an extremely 
small number count, such that the spectrum could not be calculated.

% AIM: This is the first mention of this as a classification problem. We should mention this up in experimental design along with the assumption of ML/training-based methods, but also change "clasified" [SIC] to "assigned" for overall consistency (and maybe "that entrants clasified" --> "entrants to classify," and maybe reword post-semicolon part as "aside from the impact of reduced final sample size on the metrics, no penalty was applied for discarding galaxies entirely").
% AIM: Also, the discussion of non-assignment should be its own subsection.
The challenge did not require that entrants clasified all galaxies; ambiguous
galaxies could be discarded at the cost of number density.  Several methods
took advantage of this.  The ComplexSom method (see \autoref{sec:csom}) found
it did not improve scores, but as noted in \autoref{sec:pqnld} this may be an artifact
of the high density of the training data, and so could be explored in future challenges.

@@

The overall winning method in the 3x2pt case in the fiducial $griz$ 3x2pt CosmoDC2 metric was \textsc{funbins},
which used a random forest to assign bins, but chose nominal bin edges by splitting the range spanned
by the sample into equal co-moving distance bins. As described in \autoref{sec:funbins},  this was designed
to minimze the shot noise for the angular power in each bin, and this has proven successful. 
Figure \ref{fig:funbin_nz} shows the number density obtained by this method, for both the riz and griz 
scenarios, and an additional post-challenge run using the full Rubin $ugrizy$ bands, the metric results for
which were close to those of the $griz$ analysis.

% AIM: Put the caveat at the beginning of the following sentence rather than the end. Also, this claim is not well-supported even by Fig. 3. . . 
It is clear from this that fears about restricted photometry catastrophically limiting
the information needed for tomography are unfounded, at least in our idealised scenario.
In both the CosmoDC2 and Buzzard data sets,
even the $riz$ bands alone are sufficient to split objects into nine reasonably separated
tomographic bins.  Once bins become this narrow the primary analysis concern will become the calibration 
of overall bin $n(z)$ values, rather than initial tomography, so we consider this to be reassuring.

@@

@@

\section{Mistakes} \label{sec:mistakes}
% AIM: This should not be an appendix and should instead be part of the Results & Discussion section. They should not be presented as "mistakes" -- they are part of the experimental design and constitute caveats, which are simply facts, not subjective judgements.

The challenge organizers made a number of mistakes when building and running the challenge.
These do not invalidate our process or results, but we describe some of them here in the hope of
being useful for future challenge designers.

@@

\section{Full 9-bin results table} \label{app:tables}
Tables \ref{tab:full_dc2} and \ref{tab:full_buzz} show all calculated metrics when generating nine tomographic 
bins for CosmoDC2 and Buzzard respectively. The \textit{ww} columns
show weak-lensing metrics, \textit{gg} show galaxy clustering, and \textit{3x} the full 3x2pt metric.
Methods above the horizontal line trained bin edges; methods below used fixed fiducial ones, though in some
cases did some hand-tuning of them before submission. In each section the highest scoring method is highlighted.

>>>>

This paper is part of preparations for the analysis of Legacy Survey of Space and Time (LSST) data at the Rubin Observatory, being run by the Dark Energy Science Collaboration (DESC).
In it, we discuss and evaluate the related challenge
of using a limited set of colour bands to make those assignments, motivated by requirements of using the
metacalibration method for galaxy shape measurement to limit biases.  
A similar question, with different motivations, was explored in \citet{jain}, who find that 3-band 
gri tomography could be effective for tomographic selection, a result we will echo here in a new context: in this paper we describe the results of a 
challenge using simulated Rubin-like data with a range of methods submitted by different teams.  We explore how many tomographic bins we can effectively generate using only a subset of bins, and compare methods submitted to the challenge.

@@

In the challenge, participants assigned simulated galaxies to tomographic bins,
using only their (g)riz bands and associated errors, and the galay size.  Their goal was to maximize the cosmological
information in the final split data set; this is generally acheived by separating different bins by 
redshift as much as possible.  Particpants were free to optimize or simply select nominal (target) 
redshift bin edges from their training sample, but in either case had to assign galaxies into the 
different bins.

@@

The challenge was structured as a python library, in which each assignment
method was a subclass of an abstract base parent superclass, {\sc Tomographer}.
The superclass, and other challenge machinery, performed initialization,
configuration option handling, data loading, and computing derived quantities
such as colours from magnitudes.

@@

\subsection{Results Overview}
The DETF results for each method are shown in Table \ref{tab:cosmodc2}.  Complete results for all metrics
for the nine-bin scenario only are shown in appendix tables \ref{tab:full_dc2} and \autoref{tab:full_buzz}.

In all these tables, missing entries are shown for two reasons.
The first, marked with an asterisk, is when a method did not run for the given scenario,
either due to memory overflow or taking an extremely long time.
The second, marked with a dash, is when a method ran, but generated at least one bin with an extremely 
small number count, such that the spectrum could not be calculated.

@@

The overall winning method in the 3x2pt case in the fiducial $griz$ 3x2pt CosmoDC2 metric was \textsc{funbins},
which used a random forest to assign bins, but chose nominal bin edges by splitting the range spanned
by the sample into equal co-moving distance bins. As described in \autoref{sec:funbins},  this was designed
to minimze the shot noise for the angular power in each bin, and this has proven successful. 
Figure \ref{fig:funbin_nz} shows the number density obtained by this method, for both the riz and griz 
scenarios, and an additional post-challenge run using the full Rubin $ugrizy$ bands.  The metrics
obtained with \textsc{funbins} on $ugrizy$ are barely higher than those with $griz$ -- the largest
increase is 8\% for the $\Omega_c-\sigma_8$ metric, with the remainder all 5\% or less.

It is clear from this that, at least in our idealised scenario,  fears about restricted photometry 
catastrophically limiting
the information needed for tomography are unfounded.
In both the CosmoDC2 and Buzzard data sets,
even the $riz$ bands alone are sufficient to split objects into nine reasonably separated
tomographic bins.  Once bins become this narrow the primary analysis concern will become the calibration 
of overall bin $n(z)$ values, rather than initial tomography, so we consider this to be reassuring.

@@

\subsection{Non-assignment}
The challenge did not require entrants to assign all galaxies to bins; aside from the impact of reduced final sample size on the metrics, no penalty was applied for discarding galaxies entirely.  Several methods
took advantage of this.  The ComplexSom method (see \autoref{sec:csom}) found
it did not improve scores, but as noted in \autoref{sec:pqnld} this may be an artifact
of the high density of the training data, and so could be explored in future challenges.

@@

\section{Mistakes} \label{sec:mistakes}
The challenge organizers made a number of mistakes when building and running the challenge.
These do not invalidate our process or results, but we describe some of them here in the hope of
being useful for future challenge designers.

@@

\section{Full 9-bin results tables} \label{app:tables}
Tables \ref{tab:full_dc2} and \ref{tab:full_buzz} show all calculated metrics when generating nine tomographic 
bins for CosmoDC2 and Buzzard respectively. The \textit{ww} columns
show weak-lensing metrics, \textit{gg} show galaxy clustering, and \textit{3x} the full 3x2pt metric.
Methods above the horizontal line trained bin edges; methods below used fixed fiducial ones, though in some
cases did some hand-tuning of them before submission. In each section the highest scoring method is highlighted.
